---
title: "Introduction to bulkreadr package"
output: rmarkdown::html_vignette
author: "Ezekiel Ogundepo and Ernest Fokoué"
vignette: >
  %\VignetteIndexEntry{bulkreadr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE, 
  warning = FALSE,
  comment = "#>",
  fig.path = "man/figures/",
  out.width = "100%")

options(tibble.print_min = 5, tibble.print_max = 5)

options(rmarkdown.html_vignette.check_title = FALSE)
```

## About the package

`bulkreadr` is an R package designed to simplify and streamline the process of reading and processing large volumes of data. With a collection of functions tailored for bulk data operations, the package allows users to efficiently read multiple sheets from Microsoft Excel/Google Sheets workbooks and multiple CSV files from a directory. It returns the data as organized data frames, making it convenient for further analysis and manipulation. Whether dealing with extensive data sets or batch processing tasks, "bulkreadr" empowers users to effortlessly handle data in bulk, saving time and effort in data preparation workflows.

## Installation

You can install `bulkreadr` package from [CRAN](https://cran.r-project.org/) with:

```{r eval=FALSE}
install.packages("bulkreadr")

```

or the development version from [GitHub](https://github.com/gbganalyst/bulkreadr) with


```{r eval=FALSE}

if(!require("devtools")){
 install.packages("devtools")
}

devtools::install_github("gbganalyst/bulkreadr")
```


## How to load the package

Now that you have installed `bulkreadr` package, you can simply load it by using:

```{r pkgload}
library(bulkreadr)

```

## Functions in bulkreadr package

This section provides a concise overview of the different functions available in the `bulkreadr` package. These functions serve various purposes and are designed to handle importing of data in bulk.

- [`read_excel_workbook()`](#read_excel_workbook)

- [`read_excel_files_from_dir()`](#read_csv_files_from_dir)

- [`read_csv_files_from_dir()`](#read_csv_files_from_dir)

- [`read_gsheets()`](#read_gsheets)

- [`read_spss_data()`](#read_spss_data)


## Other functions in bulkreadr package:

- [`pull_out()`](#pull_out)

- [`convert_to_date()`](#convert_to_date)

- [`inspect_na()`](#inspect_na)

- [`fill_missing_values()`](#fill_missing_values)

**Note:**

For the majority of functions within this package, we will utilize data stored in the system file by the `bulkreadr`, which can be accessed using the `system.file()` function. If you wish to utilize your own data stored in your local directory, please ensure that you have set the appropriate file path prior to using any functions provided by the bulkreadr package.


## read_excel_workbook()

`read_excel_workbook()` reads all the data from the sheets of an Excel workbook and return an appended dataframe.


```{r example1}

# path to the xls/xlsx file.

path <- system.file("extdata", "Diamonds.xlsx", package = "bulkreadr")

# read the sheets

read_excel_workbook(path = path)

```

## read_excel_files_from_dir()

`read_excel_files_from_dir()` reads all Excel workbooks in the `"~/data"` directory and returns an appended dataframe.

```{r example1a}

# path to the directory containing the xls/xlsx files.

directory <- system.file("xlsxfolder",  package = "bulkreadr")

# import the workbooks

read_excel_files_from_dir(dir_path = directory)

```

## read_csv_files_from_dir() 

`read_csv_files_from_dir()` reads all csv files from the `"~/data"` directory and returns an appended dataframe. The resulting dataframe will be in the same order as the CSV files in the directory.

```{r example2}
# path to the directory containing the CSV files.

directory <- system.file("csvfolder",  package = "bulkreadr")

# import the csv files

read_csv_files_from_dir(dir_path = directory)

```

## read_gsheets() 

The `read_gsheets()` function imports data from multiple sheets in a Google Sheets spreadsheet and appends the resulting dataframes from each sheet together to create a single dataframe. This function is a powerful tool for data analysis, as it allows you to easily combine data from multiple sheets into a single dataset.

```{r, include=FALSE}
googlesheets4::gs4_deauth()
```

```{r example3, message=FALSE, warning=FALSE}

# Google Sheet ID or the link to the sheet

sheet_id <- "1izO0mHu3L9AMySQUXGDn9GPs1n-VwGFSEoAKGhqVQh0"

# read all the sheets

read_gsheets(ss = sheet_id)
```

## read_spss_data() 

`read_spss_data()` is designed to seamlessly import data from an SPSS data (`.sav` or `.zsav`) files. It converts labelled variables into factors, a crucial step that enhances the ease of data manipulation and analysis within the R programming environment.

**Read an SPSS data file without converting variable labels as column names**

```{r, spssdata1}

file_path <- system.file("extdata", "Wages.sav", package = "bulkreadr")

data <- read_spss_data(file = file_path)

data

```


**Read an SPSS data file and convert variable labels as column names**

```{r spssdata2}

data <- read_spss_data(file = file_path, label = TRUE)

data

```

## pull_out()

`pull_out()` is similar to [. It acts on vectors, matrices, arrays and lists to extract or replace parts. It is pleasant to use with the magrittr (`⁠%>%`⁠) and base(`|>`) operators.

```{r example4}

top_10_richest_nig <- c("Aliko Dangote", "Mike Adenuga", "Femi Otedola", "Arthur Eze", "Abdulsamad Rabiu", "Cletus Ibeto", "Orji Uzor Kalu", "ABC Orjiakor", "Jimoh Ibrahim", "Tony Elumelu")

top_10_richest_nig %>% 
  pull_out(c(1, 5, 2))
```

```{r}
top_10_richest_nig %>% 
  pull_out(-c(1, 5, 2))
```


## convert_to_date()

`convert_to_date()` parses an input vector into POSIXct date-time object. It is also powerful to convert from excel date number like `42370` into date value like `2016-01-01`.

```{r example 5}

## ** heterogeneous dates **

dates <- c(
  44869, "22.09.2022", NA, "02/27/92", "01-19-2022",
  "13-01-  2022", "2023", "2023-2", 41750.2, 41751.99,
  "11 07 2023", "2023-4"
  )

# Convert to POSIXct or Date object

convert_to_date(dates)

# It can also convert date time object to date object 

convert_to_date(lubridate::now())

```

## inspect_na() 

`inspect_na()` summarizes the rate of missingness in each column of a data frame. For a grouped data frame, the rate of missingness is summarized separately for each group.

```{r example 6a}

# dataframe summary

inspect_na(airquality)
```


**Grouped dataframe summary**

```{r}
airquality %>% 
  group_by(Month) %>% 
  inspect_na()
```

## fill_missing_values() 

`fill_missing_values()` in an efficient function that addresses missing values in a dataframe. It uses imputation by function, meaning it replaces missing data in numeric variables with either the mean or the median, and in non-numeric variables with the mode. The function takes a column-based imputation approach, ensuring that replacement values are derived from the respective columns, resulting in accurate and consistent data. This method enhances the integrity of the dataset and promotes sound decision-making and analysis in data processing workflows.

```{r example 6}

df <- tibble(
  Sepal_Length = c(5.2, 5, 5.7, NA, 6.2, 6.7, 5.5),
  Sepal.Width = c(4.1, 3.6, 3, 3, 2.9, 2.5, 2.4),
  Petal_Length = c(1.5, 1.4, 4.2, 1.4, NA, 5.8, 3.7),
  Petal_Width = c(NA, 0.2, 1.2, 0.2, 1.3, 1.8, NA),
  Species = c("setosa", NA, "versicolor", "setosa",
    NA, "virginica", "setosa"
  )
)

```

```{r}
df
```


**Using mean to fill missing values for numeric variables**


```{r}

result_df_mean <- fill_missing_values(df, use_mean = TRUE)

result_df_mean

```


**Using median to fill missing values for numeric variables**

```{r}
result_df_median <- fill_missing_values(df, use_mean = FALSE)

result_df_median
```


### Impute missing values (NAs) in a grouped data frame

You can use the `fill_missing_values()` in a grouped data frame by using other grouping and map functions. Here is an example of how to do this:

```{r}
sample_iris <- tibble(
Sepal_Length = c(5.2, 5, 5.7, NA, 6.2, 6.7, 5.5),
Petal_Length = c(1.5, 1.4, 4.2, 1.4, NA, 5.8, 3.7),
Petal_Width = c(0.3, 0.2, 1.2, 0.2, 1.3, 1.8, NA),
Species = c("setosa", "setosa", "versicolor", "setosa",
          "virginica", "virginica", "setosa")
)

```

```{r}
sample_iris
```

```{r}
sample_iris %>%
  group_by(Species) %>%
  group_split() %>%
  map_df(fill_missing_values)
```

## Inspiration

bulkreadr draws on and complements / emulates other packages such as `readxl`, `readr`, and `googlesheets4` to read bulk data in R.

  * [readxl](https://readxl.tidyverse.org) is the tidyverse package for reading Excel files (xls or xlsx) into an R data frame. 
  
  * [readr](https://readr.tidyverse.org) is the tidyverse package for reading delimited files (e.g., csv or tsv) into an R data frame. 
  
  * [googlesheets4](https://cran.r-project.org/package=googlesheets) is the package to interact with Google Sheets through the Sheets API v4  <https://developers.google.com/sheets/api>.




