---
title: "Other functions in bulkreadr"
output: rmarkdown::html_vignette
author: "Ezekiel Ogundepo and Ernest Fokoué"
vignette: >
  %\VignetteIndexEntry{Other functions in bulkreadr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
description: >
  The `bulkreadr` package includes specialized functions beyond bulk data reading, aimed at enhancing data analysis efficiency. These functions are designed to operate on individual vectors, except for `inspect_na()` and `fill_missing_values()`, which work   on data frames. 
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE, 
  warning = FALSE,
  comment = "#>",
  fig.path = "man/figures/",
  out.width = "100%")

options(tibble.print_min = 5, tibble.print_max = 5)

options(rmarkdown.html_vignette.check_title = FALSE)
```

The `bulkreadr` package in R includes specialized functions beyond bulk data reading, aimed at enhancing data analysis efficiency. These functions are designed to operate on individual vectors, except for `inspect_na()` and `fill_missing_values()`, which work on data frames. 

## pull_out()

`pull_out()` is similar to [. It acts on vectors, matrices, arrays and lists to extract or replace parts. It is pleasant to use with the magrittr (`⁠%>%`⁠) and base(`|>`) operators.

```{r example4}

library(bulkreadr)
library(dplyr)

top_10_richest_nig <- c("Aliko Dangote", "Mike Adenuga", "Femi Otedola", "Arthur Eze", "Abdulsamad Rabiu", "Cletus Ibeto", "Orji Uzor Kalu", "ABC Orjiakor", "Jimoh Ibrahim", "Tony Elumelu")

top_10_richest_nig %>% 
  pull_out(c(1, 5, 2))
```

```{r}
top_10_richest_nig %>% 
  pull_out(-c(1, 5, 2))
```


## convert_to_date()

`convert_to_date()` parses an input vector into POSIXct date-time object. It is also powerful to convert from excel date number like `42370` into date value like `2016-01-01`.

```{r example 5}

## ** heterogeneous dates **

dates <- c(
  44869, "22.09.2022", NA, "02/27/92", "01-19-2022",
  "13-01-  2022", "2023", "2023-2", 41750.2, 41751.99,
  "11 07 2023", "2023-4"
  )

# Convert to POSIXct or Date object

convert_to_date(dates)

# It can also convert date time object to date object 

convert_to_date(lubridate::now())

```

## inspect_na() 

`inspect_na()` summarizes the rate of missingness in each column of a data frame. For a grouped data frame, the rate of missingness is summarized separately for each group.

```{r example 6a}

# dataframe summary

inspect_na(airquality)
```

**Grouped dataframe summary**

```{r}
airquality %>% 
  group_by(Month) %>% 
  inspect_na()
```

## fill_missing_values() 

`fill_missing_values()` is an efficient function that addresses missing values in a data frame. It uses imputation by function, also known as column-based imputation, to impute the missing values. It supports various imputation methods for continuous variables, including `minimum`, `maximum`, `mean`, `median`, `harmonic mean`, and `geometric mean`. For categorical variables, missing values are replaced with the `mode` of the column. This approach ensures accurate and consistent replacements derived from individual columns, resulting in a complete and reliable dataset for improved analysis and decision-making.


```{r example 6}

df <- tibble::tibble(
  Sepal_Length = c(5.2, 5, 5.7, NA, 6.2, 6.7, 5.5),
  Sepal.Width = c(4.1, 3.6, 3, 3, 2.9, 2.5, 2.4),
  Petal_Length = c(1.5, 1.4, 4.2, 1.4, NA, 5.8, 3.7),
  Petal_Width = c(NA, 0.2, 1.2, 0.2, 1.3, 1.8, NA),
  Species = c("setosa", NA, "versicolor", "setosa",
    NA, "virginica", "setosa"
  )
)

```

```{r}
df
```

If you do not specify `selected_variables` (i.e., leave it as `NULL`), the function will impute missing values for all columns in the dataframe.

```{r}
# Impute using the mean

result_df_mean <- fill_missing_values(df, method = "mean")

result_df_mean
```


If you specify column names, only those columns will be imputed. For example, impute for variables `Petal_Length` and `Petal_Width` using the geometric mean.

```{r}

result_df_geomean <- fill_missing_values(df, selected_variables = c
("Petal_Length", "Petal_Width"), method = "geometric")

result_df_geomean
```

If you specify column positions, only the columns at those positions will be imputed.

```{r}
# Impute using the maximum method

result_df_max <- fill_missing_values(df, selected_variables = c
(2, 3), method = "max")

result_df_geomean
```


### Impute missing values (NAs) in a grouped data frame

You can use the `fill_missing_values()` in a grouped data frame by using other 
grouping and map functions. Here is an example of how to do this:

```{r}
sample_iris <- tibble::tibble(
Sepal_Length = c(5.2, 5, 5.7, NA, 6.2, 6.7, 5.5),
Petal_Length = c(1.5, 1.4, 4.2, 1.4, NA, 5.8, 3.7),
Petal_Width = c(0.3, 0.2, 1.2, 0.2, 1.3, 1.8, NA),
Species = c("setosa", "setosa", "versicolor", "setosa",
          "virginica", "virginica", "setosa")
)

```

```{r}
sample_iris
```

```{r}
sample_iris %>%
  group_by(Species) %>%
  group_split() %>%
  map_df(fill_missing_values, method = "median")
```
